---
title: "Project 2 - Group 5 - Data 624"
output:
  html_document:
    df_print: paged
---

# Data Preprocessing

### Packages
```{r, echo=FALSE}
library(dplyr)            # data manipulation
library(plyr)             # modeling
library(knitr)            # stlying/rmd
library(kableExtra)       # stlying
library(ggplot2)          # visualization
library(purrr)            #
library(tidyr)            #
library(corrplot)         # correlation plot
library(MASS)             #
library(caret)            # modeling
library(e1071)            # modeling
library(ranger)           # modeling
library(xgboost)          # modeling
```

### Load Data
```{r}
data = read.csv("https://raw.githubusercontent.com/hvasquez81/Data-624/master/Project%202/StudentData%20-%20TO%20MODEL.csv")
str(data) 
```

### Summary of Data
```{r}
#remove records where pH is null
data = data %>%
  filter(!is.na(PH))
```

# NAs
```{r}
#check for NA's and distribution of vars
NA_summary = data %>%
  summarise_all( funs( round( sum( 100*is.na(.)/nrow(data) ), 2) ) ) %>%
  t()
NA_summary = cbind(as.character( rownames(NA_summary) ), data.frame(NA_summary, row.names=NULL))
colnames(NA_summary) = c("Variable", "Percent of NAs")
NA_summary %>%
  arrange( desc(`Percent of NAs`) ) %>%
  head(10) %>%
  kable() %>%
  kable_styling()
```

### 0's
```{r}
data.zero = data %>%
  group_by(.) %>%
  summarise_each(funs(sum(.==0, na.rm = TRUE))) %>%
  t() %>%
  as.data.frame()

data.zero = as.data.frame(cbind(rownames(data.zero), data.zero))
rownames(data.zero) = c()
colnames(data.zero) = c('Variable', 'Zero Count')

data.zero %>% 
  arrange(desc(`Zero Count`)) %>%
  head(5) %>%
  ggplot(aes(x=Variable, y=`Zero Count`, fill = Variable)) + 
  geom_bar(stat="identity") +
  theme_classic()

```

### Create new variable HydPressureRecorded123 
```{r}
data = data %>%
  mutate(HydPressureRecorded123 =ifelse( ( (!is.na(Hyd.Pressure1) & Hyd.Pressure1 !=0 ) |
                                            (!is.na(Hyd.Pressure2) & Hyd.Pressure2 !=0 ) |
                                              (!is.na(Hyd.Pressure3) & Hyd.Pressure3 !=0 ) ), 1, 0)
  )
#change to factor
data$HydPressureRecorded123 = as.factor(data$HydPressureRecorded123)
```

### Variable distribution
```{r}
data %>%
  keep(is.numeric) %>%
  gather() %>%
  ggplot(aes(value)) + 
  facet_wrap(~ key, scales = "free") + 
  geom_histogram(aes(fill=key)) + 
  theme(legend.position = "none") +
  theme_classic()
    
```

### Categorical Variable Distribution
```{r}
#replace blank brand code with E for empty
data$Brand.Code = as.character(data$Brand.Code)
data$Brand.Code[is.na(data$Brand.Code)] = "E"
data$Brand.Code[data$Brand.Code == ""] = "E"
data$Brand.Code = as.factor(data$Brand.Code)

brand_cnt = data %>% 
  group_by(Brand.Code) %>% 
  tally()

ggplot(brand_cnt, aes(x=Brand.Code, y=n, fill = Brand.Code)) + 
  geom_bar(stat="identity") +
  theme_classic()
```

```{r}
HydPressureRecorded123_cnt = data %>% 
  group_by(HydPressureRecorded123) %>% 
  tally()

ggplot(HydPressureRecorded123_cnt, aes(x=HydPressureRecorded123, y=n, fill = HydPressureRecorded123)) + 
  geom_bar(stat="identity") +
  theme_classic()
```

### Correlation Matrix
```{r}
#use numeric variables
numeric_data = data %>%
  keep(is.numeric)
#create correlation matrix
corrplot(cor(numeric_data, use = "complete.obs"),
         type = "lower",
         method = "circle")
```

# Modeling

## Data imputation
```{r}
#replace na function
NA_replace <- function(x) replace(x, is.na(x), mean(x, na.rm = TRUE))


#removes categorical variables and then replaces NAs of numeric variables with column means
data[, !( colnames(data) %in% c('Brand.Code', 'HydPressureRecorded123') ) ] = replace(data[, !( colnames(data) %in% c('Brand.Code', 'HydPressureRecorded123') ) ], TRUE, lapply(data[, !( colnames(data) %in% c('Brand.Code', 'HydPressureRecorded123') ) ], NA_replace ))
  
```

## Train Test Split
```{r}
train_size = floor(0.8*nrow(data))

set.seed(1234)
train_rows = sample(seq_len(nrow(data)), size = train_size)
train = data[train_rows, ]
test = data[-train_rows, ]
```

## Linear

### Multiple Linear Regression Model
```{r}
set.seed(1234)
current.time = proc.time()
linear.mod = caret::train(PH~., 
                          data = train, 
                          method = "lm",
                          preProc = c("BoxCox"))
lm_train.time = proc.time() - current.time
summary(linear.mod)
```

#### Residuals for lm
```{r}
par(mfrow = c(2,2))
plot(linear.mod$finalModel)
```

## Non-Linear Models

### SVM
```{r}
set.seed(1234)
current.time = proc.time()
svm.mod = caret::train(PH~., 
                       data = train, 
                       method="svmRadial", 
                       preProc=c("center", "scale"),
                       trControl=trainControl(method="repeatedcv", number=10, repeats=3),
                       tuneLength = 10
                       )
svm_train.time = proc.time() - current.time
svm.mod
```

#### RMSE vs Cost
```{r}
plot(svm.mod)
```

### kNN
```{r}
set.seed(1234)
current.time = proc.time()
knn.mod = caret::train(PH~., 
                       data = train, 
                       method="knn",
                       preProc=c("center", "scale"),
                       trControl=trainControl(method="repeatedcv", number=10, repeats=3)
                       )
knn_train.time = proc.time() - current.time
knn.mod
```


```{r}
plot(knn.mod)
```

## Trees

### Random Forest
```{r}
set.seed(1234)
current.time = proc.time()
rf.mod = caret::train(PH~., 
                          data = train, 
                          method = "rf" ,
                          preProc=c("center", "scale"),
                          trControl=trainControl(method="repeatedcv", number=10, repeats=3)
                          )
rf_train.time = proc.time() - current.time
rf.mod
```

```{r}
plot(rf.mod)
```

### Ranger
```{r}
set.seed(1234)
current.time = proc.time()
ranger.mod = caret::train(PH~., 
                          data = train, 
                          method = "ranger" ,
                          preProc=c("center", "scale"),
                          importance = "impurity",
                          trControl=trainControl(method="repeatedcv", number=10, repeats=3)
                          )
ranger_train.time = proc.time() - current.time
ranger.mod
```

```{r}
plot(ranger.mod)
```

### Boost
```{r,warning=FALSE}
set.seed(1234)
current.time = proc.time()
boost.mod = caret::train(PH~., 
                          data = train, 
                          method = "xgbTree",
                          preProc=c("center", "scale"),
                          importance = "impurity",
                          trControl=trainControl(method="repeatedcv", number=10, repeats=3)
                          )
boost_train.time = proc.time() - current.time
boost.mod
```

```{r}
plot(boost.mod)
```

# Test Models

### Time to Train
```{r}
model_names = as.data.frame( c("lm", "SVM", "KNN", "Ranger", "rf", "Boost") )
model_train_times = as.data.frame( c(lm_train.time[3], 
                                     svm_train.time[3], 
                                     knn_train.time[3], 
                                     ranger_train.time[3], 
                                     rf_train.time[3], 
                                     boost_train.time[3] 
                                     )
                                   )
model_train.times = cbind( model_names, model_train_times)
colnames(model_train.times) = c("Model", "Elapsed Time")

#plot
ggplot( model_train.times, aes(x=Model, y=`Elapsed Time`, fill = Model) ) + 
  geom_bar(stat="identity") +
  coord_flip() +
  theme_classic()
```

### RMSE, RSquared, MAE
```{r}
# RMSE
#lm
lm.score = postResample(predict(linear.mod, newdata = test), test$PH)

#SVM
SVM.score = postResample(predict(svm.mod, newdata = test), test$PH)

#KNN 
KNN.score = postResample(predict(knn.mod, newdata = test), test$PH)

#rf
rf.score = postResample(predict(rf.mod, newdata = test), test$PH)

#Ranger
ranger.score = postResample(predict(ranger.mod, newdata = test), test$PH)

#Boost
boost.score = postResample(predict(boost.mod, newdata = test), test$PH)

#r bind RMSE, R2 and MAE
models.metrics = cbind( as.data.frame( c("lm", "SVM", "KNN", "Ranger", "rf", "Boost") ), 
       as.data.frame( rbind(lm.score, SVM.score, KNN.score, ranger.score, rf.score, boost.score) ),
       row.names = NULL
       )
colnames(models.metrics) = c("Model", "RMSE", "RSquared", "MAE")
```

### Display - RMSE
```{r}
test_rmse = ggplot( models.metrics, aes(x=Model, y=RMSE, fill = Model) ) + 
  geom_bar(stat="identity") +
  ggtitle("RMSE by Model") +
  geom_text( aes( label=round(RMSE,4) ), vjust=-0.5 ) +
  theme_classic()
test_rmse
```

#Display - RSquared
```{r}
test_rsquared = ggplot( models.metrics, aes(x=Model, y=RSquared, fill = Model) ) + 
  geom_bar(stat="identity") +
  ggtitle("R-Squared by Model") +
  geom_text( aes( label=round(RSquared,4) ), vjust=-0.5 ) +
  theme_classic()
test_rsquared
```

# Display - MAE
```{r}
test_mae = ggplot( models.metrics, aes(x=Model, y=MAE, fill = Model) ) + 
  geom_bar(stat="identity") +
  ggtitle("MAE by Model") +
  geom_text( aes( label=round(MAE,4) ), vjust=-0.5 ) +
  theme_classic()
test_mae
```

### Variable Importance - lm, kNN, SVM and Ranger
```{r}
par(mfrow=c(2,2))
plot(varImp(linear.mod), top=10, main="lm")
plot(varImp(knn.mod), top=10, main="kNN")
plot(varImp(svm.mod), top=10, main="SVM")
plot(varImp(ranger.mod), top=10, main="Ranger")
plot(varImp(rf.mod), top=10, main="rf")
plot(varImp(boost.mod), top=10, main="Boost")

```

### Predict vs Actual Values
```{r}
#create table
pred_actual = cbind( data.frame(predict(ranger.mod, newdata = test), row.names = NULL),
       test$PH
)
colnames(pred_actual) = c("Predicted", "Actual")
#MAE field
pred_actual = pred_actual %>%
  mutate(`Absolute Error` = abs(Predicted - Actual))

#plot
ggplot(pred_actual) + 
  geom_point( aes(x=Predicted, y=Actual, color = `Absolute Error`) ) +
  ggtitle("Ranger: Actual vs Predicted") +
  theme_classic()
```


```{r}
ranger.mod$finalModel
```

# Predictions on Eval Dataset

### Load Eval Data
```{r}
# load data
eval = read.csv("https://raw.githubusercontent.com/hvasquez81/Data-624/master/Project%202/StudentEvaluation-%20TO%20PREDICT.csv")
head(eval)
```

### Eval Data Manipulation and Preprocessing
```{r}
#drop the pH column
eval = eval %>%
  dplyr::select(-PH)
#replace blank brand code with E for empty
eval$Brand.Code = as.character(eval$Brand.Code)
eval$Brand.Code[is.na(eval$Brand.Code)] = "E"
eval$Brand.Code[eval$Brand.Code == ""] = "E"
eval$Brand.Code = as.factor(eval$Brand.Code)

#create new variable
eval = eval %>%
  mutate(HydPressureRecorded123 =ifelse( ( (!is.na(Hyd.Pressure1) & Hyd.Pressure1 !=0 ) |
                                            (!is.na(Hyd.Pressure2) & Hyd.Pressure2 !=0 ) |
                                              (!is.na(Hyd.Pressure3) & Hyd.Pressure3 !=0 ) ), 1, 0)
  )

#change to factor
eval$HydPressureRecorded123 = as.factor(eval$HydPressureRecorded123)

#apply replace NA function on non-categorical variables
eval[, !( colnames(eval) %in% c('Brand.Code', 'HydPressureRecorded123') ) ] = replace(eval[, !( colnames(eval) %in% c('Brand.Code', 'HydPressureRecorded123') ) ], TRUE, lapply(eval[, !( colnames(eval) %in% c('Brand.Code', 'HydPressureRecorded123') ) ], NA_replace ))
head(eval)
```

### Use Ranger to Predict on Eval set
```{r}
#store the predictions as a new row named PH_RANGER_PREDICTION and add the original PH column back
eval = eval %>% mutate(PH_RANGER_PREDICTION = predict(ranger.mod, newdata = eval),
                PH = NA)

#write file as CSV to send off
#write.csv(eval, "Group5_pH_Eval_Predictions.csv")
```

# Sources
https://rdrr.io/cran/caret/man/models.html
http://www.rebeccabarter.com/blog/2017-11-17-caret_tutorial/
https://cran.r-projecZt.org/web/packages/ranger/ranger.pdf
https://cran.r-project.org/web/packages/xgboost/vignettes/xgboost.pdf


